---
title: "SC_workflow"
author: "Stefania Pirrotta"
date: "2026-06-18"
output: html_document
---

# Prepare Data
In this workflow, we present the analyses to investigate with `mitology` the scRNAseq dataset of Papa et al. (GEO ID: GSE110973, Papa L, Zimran E, Djedaini M et al. Ex vivo human HSC expansion requires coordination of cellular reprogramming with mitochondrial remodeling and p53 activation. Blood Advances, 2018;2(20):2766–2779). We download the dataset using the `GEOquery` R package (version 2.80.0). The `SingleCellExperiment` object provides the raw count expression values.

```{r}
library(GEOquery)
filePaths = getGEOSuppFiles("GSE110973")

library(Matrix)
library(rhdf5)

data_path <- "GSE110973/GSE110973_RAW/"

h5_file_0 <- paste0(data_path, "GSM3019610_day0_gene_bc_matrices.h5")
h5_0 <- h5read(h5_file_0, "hg19")
counts_0 <- sparseMatrix(
  dims = h5_0$shape, i = as.numeric(h5_0$indices), index1 = FALSE,
  p = as.numeric(h5_0$indptr), x = as.numeric(h5_0$data))
colnames(counts_0) <- paste(h5_0$barcodes, "0", sep = "_")
rownames(counts_0) <- h5_0$genes

h5_file_2 <- paste0(data_path, "GSM3019611_day2_gene_bc_matrices.h5")
h5_2 <- h5read(h5_file_2, "hg19")
counts_2 <- sparseMatrix(
  dims = h5_2$shape, i = as.numeric(h5_2$indices), index1 = FALSE,
  p = as.numeric(h5_2$indptr), x = as.numeric(h5_2$data))
colnames(counts_2) <- paste(h5_2$barcodes, "2", sep = "_")
rownames(counts_2) <- h5_2$genes

counts <- cbind(counts_0, counts_2)

library(SingleCellExperiment)
sce <- SingleCellExperiment(
    list(counts = counts),
    colData = DataFrame(day = c(rep("d0", length(h5_0$barcodes)),
                                rep("d2", length(h5_2$barcodes)))),
    rowData = DataFrame(SYMBOL = h5_0$gene_names),
    metadata=list(study = "GSE110973"))
```

We then normalized (scuttle, v 1.18.0) and corrected data for the batch effect (batchelor, v 1.24.0).

```{r}
library(scuttle)
sce <- logNormCounts(sce)
set.seed(12345)
dec <- scran::modelGeneVar(sce, block = sce$day)
chosen.hvgs <- dec$bio > 0

library(scater)
sce <- runPCA(sce, subset_row = chosen.hvgs, ntop = 1000)
sce <- runTSNE(sce, dimred = "PCA")

# Batch correction
sce <- sce[rowSums(assay(sce, 2)) > 0,]
dec <- scran::modelGeneVar(sce, block = sce$day)
chosen.hvgs <- dec$bio > 0
assay(sce, i = 1) <- as.matrix(assay(sce, i = 1))
assay(sce, i = 2) <- as.matrix(assay(sce, i = 2))

library(batchelor)
merged <- correctExperiments(
  sce, batch = sce$day, subset.row = chosen.hvgs, correct.all = TRUE, 
  PARAM = ClassicMnnParam(BPPARAM = BiocParallel::MulticoreParam(workers = 20)))
```

We performed t-SNE using the top 50 components of the principal component analysis (scater, v 1.40.0).

```{r}
merged <- runPCA(merged, subset_row = chosen.hvgs, ntop = 1000, exprs_values = "corrected")
merged <- runTSNE(merged, dimred = "PCA")

# Figure 1E
plotTSNE(merged, colour_by = "day", point_alpha = 0.4, point_size = 1.5) +
    ggtitle("TSNE of samples batch corrected") +
    theme(axis.text = element_text(size = 12),
          axis.title = element_text(size = 12, face = "bold"), 
          title = element_text(size = 12, face = "bold"), 
          legend.text = element_text(size = 10))
```

Then, we found marker genes of cells in d0 and d2 using the SingleR package (v 2.10.0).

```{r}
genes <- SingleR::getClassicMarkers(ref = merged, labels = merged$day, assay.type = "corrected")
genes0 <- mapIds(x = org.Hs.eg.db, keys = genes$d0$d2, column = "ENTREZID", keytype = "ENSEMBL")
genes2 <- mapIds(x = org.Hs.eg.db, keys = genes$d2$d0, column = "ENTREZID", keytype = "ENSEMBL")
```

# Enrichment Analysis
We then computed two different types of enrichment analysis on these data: (i) the single-sample gene set enrichment analysis (ssGSEA) scores on the MitoCarta3.0 pathways

```{r}
library(mitology)
library(GSVA)

MC_list <- getGeneSets(database = "MitoCarta")
ssgseaMC <- gsva(ssgseaParam(exprData = merged, assay = "corrected", geneSets = MC_list))
ssgseaMCd <- do.call(cbind, lapply(unique(merged$day), function(x){
  rowMedians(assay(ssgseaMC)[,merged$day==x]) }))
colnames(ssgseaMCd) <- unique(merged$day)
rownames(ssgseaMCd) <- rownames(ssgseaMC)

# Figure 1F
mitoTreeHeatmap(data = ssgseaMCd, database = "MitoCarta", labelNames = "sections", font.size = 3)

colData(merged) <- cbind(colData(merged), OXPHOS_subunits = assay(ssgseaMC)["OXPHOS_subunits",])
plotTSNE(merged, point_alpha = 0.4, point_size = 1, colour_by = "OXPHOS_subunits") +
    theme(legend.position = "top") +
  scale_color_gradient2(low = c("#701d46", "#CB347F"), mid = "white", high = plotPar$DBcol)

colData(merged) <- cbind(
    colData(merged), Glycerol_phosphate_shuttle = assay(ssgseaMC)["Glycerol_phosphate_shuttle",])
plotTSNE(merged, point_alpha = 0.4, point_size = 1, colour_by = "Glycerol_phosphate_shuttle") +
    theme(legend.position = "top") +
  scale_color_gradient2(low = c("#701d46", "#CB347F"), mid = "white", high = plotPar$DBcol)

colData(merged) <- cbind(
  colData(merged), ROS_and_glutathione_metabolism = assay(ssgseaMC)["ROS_and_glutathione_metabolism",])
plotTSNE(merged, point_alpha = 0.4, point_size = 1, colour_by = "ROS_and_glutathione_metabolism") +
    theme(legend.position = "top") +
  scale_color_gradient2(low = c("#701d46", "#CB347F"), mid = "white", high = plotPar$DBcol)
```


and (ii) the over-representation analysis (ORA) on the Reactome gene sets in mitology and the original Reactome pathways.

```{r}
orRT0 <- enrichMito(genes = genes$d0$d2, database = "Reactome")
orRT2 <- enrichMito(genes = genes$d2$d0, database = "Reactome")

# Figure 1G
mitoTreePoint(data = list(
  "mitology_d0" = orRT0$mitology, "mitology_d2" = orRT2$mitology, 
  "original_d0" = orRT0$original, "original_d2" = orRT2$original), 
              database = "Reactome", pvalCutoff = .1) + ggtitle("Reactome")
```

