---
title: "ST_workflow"
author: "Stefania Pirrotta"
date: "2026-06-18"
output: html_document
---

# Prepare Data
In this workflow, we present the analyses to investigate with `mitology` the ST dataset of Lin et al. (GEO ID: GSE284089, Lin W, Li Y, Qiu C et al. Mapping the spatial atlas of the human bone tissue integrating spatial and single-cell transcriptomics. Nucleic Acids Research, 2025;53(2):gkae1298). We download the dataset using the `GEOquery` R package (version 2.80.0). The `SpatialExperiment` object provides the raw count expression values.

```{r}
library(GEOquery)
gse <- getGEO("GSE284089", GSEMatrix = TRUE, AnnotGPL = TRUE, destdir = dir)[[1]]
metadata <- pData(gse)
getGEOSuppFiles("GSE284089", baseDir = dir)

library(SpatialExperiment)
sce <- DropletUtils::read10xCounts("GSE284089/GSE284089_RAW_spe/raw_feature_bc_matrix")
img <- readImgData(path = "GSE284089/GSE284089_RAW_spe/spatial", sample_id = "GSE284089")
xyz <- read.csv("GSE284089/GSE284089_RAW_spe/spatial/tissue_positions.csv.gz", header = TRUE)
xyz <- xyz[match(sce@colData@listData[["Barcode"]], xyz$barcode),]
rd <- S4Vectors::DataFrame(symbol = rowData(sce)$Symbol)

spe <- SpatialExperiment(
  assays = list(counts = assay(sce)),
  rowData = rd, 
  colData = DataFrame(xyz), 
  spatialCoordsNames = c("pxl_col_in_fullres", "pxl_row_in_fullres"),
  imgData = img,
  sample_id = "GSE284089")
```

We then performed the quality control (scater, v 1.40.0), normalization (scran, v 1.40.0), feature selection and dimensionality reduction.

```{r}
library(scater)
spe <- addPerCellQC(spe)
spe <- spe[, spe$sum >= 100]

library(scran)
spe <- spe[apply(assay(spe) > 0, 1, any),]
spe <- computeLibraryFactors(spe)
spe <- logNormCounts(spe)

dec <- modelGeneVar(spe)
hvgs <- getTopHVGs(dec, prop = .2)

set.seed(12345)
spe <- runPCA(spe, subset_row = hvgs)
spe <- runUMAP(spe, dimred = "PCA")
```

We performed Leiden clustering using the components of the principal component analysis.

```{r}
k <- 50
g <- buildSNNGraph(spe, k = k, use.dimred = "PCA")
k <- igraph::cluster_leiden(g, objective_function="modularity", resolution=.6)
spe$Leiden <- factor(k$membership)

# Figure 1H
plotVisium(spe, annotate = "Leiden", image = TRUE)
```

# Enrichment Analysis
We then computed the ssGSEA scores on Reactome database included in mitology and the corresponding original database.

```{r}
library(mitology)
library(GSVA)
RT_list <- getGeneSets(database = "Reactome")
ssgseaRes <- gsva(ssgseaParam(exprData = spe, assay = "logcounts", geneSets = RT_list))
ssgseaResA <- aggregateAcrossCells(
    ssgseaRes, ids = spe$Leiden, use.assay.type = "es", statistics = "mean")

library(graphite)
humanReactome <- pathways("hsapiens", "reactome")
hREnsembl <- convertIdentifiers(humanReactome, "ENSEMBL")
hREnsembl_g <- lapply(hREnsembl, function(x){substring(nodes(x),9)})
hREnsembl_g <- hREnsembl_g[lengths(hREnsembl_g)>0]
hREnsembl_g <- hREnsembl_g[!duplicated(names(hREnsembl_g))]
names(hREnsembl_g) <- mitology:::.mitologyName(names = names(hREnsembl_g), database = "Reactome")
hREnsembl_g <- hREnsembl_g[names(hREnsembl_g) %in% names(RT_list)]
ssgseaRTall <- gsva(ssgseaParam(
    exprData = spe, assay = "logcounts", geneSets = hREnsembl_g, minSize = 10))
ssgseaRTallA <- aggregateAcrossCells(
  ssgseaRTall, ids = spe$Leiden, use.assay.type = "es", statistics = "mean")

pvals <- list()
for (db in c("RT", "RTall")){
  pvals_db <- c()
  if (db == "RT"){ dbes <- ssgseaRes } else { dbes <- ssgseaRTall }
  for (i in 1:nrow(assay(dbes))){
    t_ssgsea <- t.test(as.vector(assay(dbes)[i, spe$Leiden==1]), 
                       as.vector(assay(dbes)[i, spe$Leiden==2]))
    pvals_db <- c(pvals_db, t_ssgsea$p.value) }
  names(pvals_db) <- rownames(dbes)
  pvals[[db]] <- pvals_db }

pvals <- lapply(pvals, method = "fdr", p.adjust)
pvals <- lapply(pvals, function(x) x[x < .0001])

ssgseaRTA <- merge(assay(ssgseaResA[names(pvals$RT),]), 
                   assay(ssgseaRTallA[names(pvals$RTall),]), 
                   by = 0, all = TRUE)
rownames(ssgseaRTA) <- ssgseaRTA$Row.names
ssgseaRTA <- ssgseaRTA[,2:5]
colnames(ssgseaRTA) <- c("n1_mito", "n2_mito", "n1_orig", "n2_orig")

# Figure 1I
mitoTreeHeatmap(data = ssgseaRTA, database = "Reactome", labelNames = "leaves", font.size = 3)
```

